---
title: "WindmillDepth"
author: "Silas Christian Wilhelm-Leth"
date: "June 2025"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# OBJECTIVE

In this script, I will be analyzing several data types related to offshore windmills, including: depth data (as bathymetry raster), point data of existing turbines, shapefile of protected natura 2000 areas, shapefile shipping lane data, and raster wind power data.

# Lets get started

Let's load our libraries. 

```{r}
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
```

I'll load the bathymetry raster and the windmill points
```{r}
depth_raster <- rast("data/DK_Dybdemodel_2024/ddm_50m.dybde.tiff")
windmills <- st_read("data/ens_vindkraftanlaeg.gpkg")

#And let's see in what layer the points are
st_layers("data/ens_vindkraftanlaeg.gpkg")
```

# Cleanup

I only want the layer with points, therefore I'll use st_read to create my "windmills" object again, with only what I need
```{r}
windmills <- st_read("data/ens_vindkraftanlaeg.gpkg", layer = "vindkraftanlaeg_historisk_indevaerende")
```

Next: Using st_transform, I'll make sure that the windmill points are in the same crs as the depth_raster
```{r}
windmills <- st_transform(windmills, crs(depth_raster))
```


Before moving on, I need to check the crs of my data. I can see that my transform worked and they are the same.

```{r}
crs(depth_raster) # It's in 3034, as is seen in the bottom right
st_crs(windmills) # It's at the very bottom, also 3034

```
Let's test that everything works with the windmills...

```{r}
nrow(windmills)
head(windmills)
plot(st_geometry(windmills))

```
That the bounding boxes fit...

```{r}
ext(depth_raster)  
st_bbox(windmills)

```

And for good measure, a visual sanity check, confirming that the bathymetry raster and the points work with each other

```{r}
plot(depth_raster, main = "Bathymetry with windmills")
plot(st_geometry(windmills), add = TRUE, col = "red", pch = 16)

```

# Final part of cleanup

As I am working with offshore windmills only, we need to remove all the onshore ones. In other words, removing the points which fall outside the bathymetry raster. I attempted to do this with a polygon mask and intersect, but I did not have a polygon, and creating one proved difficult to do, without also losing important data. Instead, I will simply remove all windmills which are placed at NA points on the bathymetry raster. (meaning they are on land)
This less direct method is not as generally applicable - only made possible by the specific nature of this data. It also won't allow for easy reuse and further comparisons, to the same extent as a polygon/mask would have.

First, we get the depth values via terra extract.

```{r}
depth_values <- terra::extract(depth_raster, vect(windmills))
summary(depth_values[, 2])

```

This shows me, that the depth values are measured as positive numbers. This won't do any difference for the analysis - just something to note.

```{r}
sum(!is.na(depth_values[, 2]))
sum(depth_values[, 2] < 0, na.rm = TRUE)

```
Filtering out the onshore windmills:

```{r}
windmills$depth <- depth_values[, 2]

plot(depth_raster, main = "Windmills - offshore only")
plot(st_geometry(windmills), col = ifelse(is.na(windmills$depth), "gray", 
                                          ifelse(windmills$depth >= 0, "red", "gray")),
     pch = 16, add = TRUE)

legend("bottomleft", legend = c("Offshore", "Onshore (no raster data)"),
       col = c("red", "gray"), pch = 16, bty = "n")

```

Cleaning up the onshore windmills.

```{r}
# Step 1: Extract bathymetry values at windmill locations
depth_values <- terra::extract(depth_raster, vect(windmills))

# Step 2: Add extracted values as a new column
windmills$depth <- depth_values[, 2]

# Step 3: Keep only windmills placed over water (i.e., in raster coverage)
# Optional: you can include '& depth > 0' if needed
offshore_windmills <- windmills[!is.na(windmills$depth), ]

```

The clean result:

```{r}
plot(depth_raster, main = "Offshore windmills on bathymetry")
plot(st_geometry(offshore_windmills), add = TRUE, col = "red", pch = 16)

```

#1# Analysis Part 1 - Depth

Let's start by calculating the mean depth at which windmills are placed. We've separated our offshore windmills into the "offshore_windmills" dataset, which includes a "depth" column. 

```{r}
mean_depth <- mean(offshore_windmills$depth, na.rm = TRUE)

# I'll print the resulting value
cat("Average offshore wind turbine depth:", round(mean_depth, 2), "meters\n")

```

```{r}
# Basic summary
summary(offshore_windmills$depth)

# Min and max
min_depth <- min(offshore_windmills$depth, na.rm = TRUE)
max_depth <- max(offshore_windmills$depth, na.rm = TRUE)

cat("Depth range:", round(min_depth, 2), "to", round(max_depth, 2), "meters\n")


```

```{r}
hist(offshore_windmills$depth,
     breaks = 30,
     col = "seagreen",
     main = "Foundation depths of offshore windmills",
     xlab = "Depth (meters below surface)")

```

```{r}
ggplot(offshore_windmills) +
  geom_sf(aes(color = depth)) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Offshore Wind Turbines by Depth",
       color = "Depth (m)")

```


```{r}
# Creating a binary raster: 1 if within depth range, NA otherwise
possible_depth <- classify(depth_raster, matrix(c(-Inf, 5, NA, 
                                                   5, 30, 5, 
                                                  30, Inf, NA), 
                                                 ncol = 3, byrow = TRUE))

# plotting this raster
# Starting with the base info and depth raster in the background
plot(depth_raster, col = gray.colors(100), main = "Possible depths (5–31 meters)")

# Adding the possible areas in color on top
plot(possible_depth, col = "seagreen3", add = TRUE, legend = FALSE)

```

#Protected Areas


```{r}
NaturaAreas <- st_read("data/np3_2022_N2000plan3_endelig_shp/np3_2022_natura_2000_omraader.shp")
```
I'll plot it next:

```{r}
plot(NaturaAreas, max.plot = 10)
```

There's clearly a lot of attributes. I'm just interested in the area. Let's map this by itself. 
Note: The attribute is seemingly colored according to size of contiguous or themed areas. We can just ignore this. 

```{r}
plot(NaturaAreas["Shape_Area"], main = "Natura 2000 areas")

```

Before the data can work together, we need to check and align the crs as before:

```{r}
crs(depth_raster)
st_crs(NaturaAreas)
```
And as earlier, we can reproject the Natura Areas crs into 3034 (from 25832), in order to match our raster

```{r}
NaturaAreas <- st_transform(NaturaAreas, crs = crs(depth_raster))
```

Checking that it worked:
```{r}
st_crs(NaturaAreas)
```

```{r}
NaturaAreas_vect <- vect(NaturaAreas)
```


```{r}
possible_no_natura <- mask(possible_depth, NaturaAreas_vect, inverse = TRUE)
```

```{r}
plot(depth_raster, col = gray.colors(100), main = "5-30 meter depths (excluding Natura 2000 zones)")

plot(possible_no_natura, col = "seagreen3", add = TRUE)
```

# Depth zones by cost

Let's make this more detailed and nuanced. I'll split depths into four zones, outlined in a 2009 report by the European Environmental Agency. 

```{r}
# Define depth classes: [min, max, value]
depth_classes <- matrix(c(
  -Inf, 10, NA,       # Ignore shallow water
   10, 20, 1,
   20, 30, 2,
   30, 40, 3,
   40, 50, 4,
   50, Inf, NA        # Ignore too deep
), ncol = 3, byrow = TRUE)

# Apply classification
depth_zones <- classify(depth_raster, rcl = depth_classes)
```

```{r}
# Reuse your projected Natura 2000 vect
depth_zones_clean <- mask(depth_zones, NaturaAreas_vect, inverse = TRUE)

```

```{r}
# Define colors and labels
zone_colors <- c("skyblue1", "deepskyblue3", "slateblue3", "darkslateblue")
zone_labels <- c("10–20m", "20–30m", "30–40m", "40–50m")

# Plot it
plot(depth_zones_clean,
     col = zone_colors,
     legend = FALSE,
     main = "Depth-Based Offshore Zones (Excluding Natura 2000)")

# Add legend manually
legend("bottomleft", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")

```
Version overlaid onto the bathymetry:

```{r}
plot(depth_raster, col = gray.colors(100))

plot(depth_zones_clean,
     col = zone_colors,
     legend = FALSE,
     add = TRUE,
     main = "Depth-Based Offshore Zones (Excluding Natura 2000)")

legend("right", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")
```

# Wake effect exclusion zones - buffers around existing turbines

```{r}
offshore_windmills_proj <- st_transform(windmills, crs = 3034)

# Buffer each turbine by 5.5 km
wake_buffers <- st_buffer(offshore_windmills_proj, dist = 5500)  # meters

# Merge overlapping buffers into one shape
wake_area <- st_union(wake_buffers)

```

```{r}
# Convert to terra format
wake_vect <- vect(wake_area)

# Mask these zones out
depth_zones_final <- mask(depth_zones_clean, wake_vect, inverse = TRUE)

```
```{r}
plot(depth_zones_final,
     col = zone_colors,
     legend = FALSE,
     main = "Final Zones (Excludes Natura 2000 + Wake Buffers)")
legend("bottomleft", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")

```

```{r}
plot(depth_raster, col = gray.colors(100),
     legend = FALSE)

plot(depth_zones_final,
     col = zone_colors,
     legend = FALSE,
     add = TRUE,
     main = "Final Zones (Excludes Natura 2000 + Wake Buffers)")
legend("topright", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")
```

# Shipping lanes


```{r}
# Load the lines
shipping_routes <- st_read("data/shipping_routes_3034.shp")

# Ensure it's in the correct CRS (should already be EPSG:3034)
st_crs(shipping_routes)

# Create buffer (e.g., 2.5 km each side of the route = 5 km total width)
shipping_buffer <- st_buffer(shipping_routes, dist = 4600)

# Convert to terra format if needed
shipping_vect <- terra::vect(shipping_buffer)

```



```{r}
# Remove these from the depth zones
zones3 <- mask(depth_zones_final, shipping_vect, inverse = TRUE)

# Plot
plot(zones3,
     col = zone_colors,
     legend = FALSE,
     main = "Zones excluding Natura 2000 + Wake + Shipping Routes")
legend("bottomleft", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")

```

# Wind Power

```{r}
# Ensure both are loaded
wind_power <- rast("data/DNK_power-density_100m.tif")        # wind power
zones3      # your final suitability zones raster (already loaded)

# Reproject wind power raster to match CRS, resolution, and extent of zones3
wind_power_aligned <- project(wind_power, zones3, method = "bilinear")

# Now mask it by the final suitable zones
wind_suitability <- mask(wind_power_aligned, zones3)

# Plot result
plot(wind_suitability, main = "Wind Power Density in Suitable Zones")

```
# Final, cleaned plots

```{r}
par(mfrow = c(1, 2))  # side-by-side plots

plot(wind_suitability,
     col = hcl.colors(50, "YlOrRd", rev = TRUE),
     main = "Wind Power Density")

plot(zones3,
     col = zone_colors,
     legend = FALSE,
     main = "Depth-Based Suitability")
legend("bottomright", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")

par(mfrow = c(1, 1))  # reset

```


```{r}
# Create semi-transparent gray colors for the bathymetry
gray_transparent <- adjustcolor(gray.colors(100), alpha.f = 0.3)

# Plot with transparency in background
par(mfrow = c(1, 2))

# Wind Power
plot(depth_raster, col = gray_transparent, legend = FALSE, main = "")
plot(wind_suitability, col = hcl.colors(50, "YlOrRd", rev = TRUE),
     legend = FALSE, add = TRUE, main = "Wind Power Density")

# Depth Suitability
plot(depth_raster, col = gray_transparent, legend = FALSE, main = "")
plot(zones3, col = zone_colors, legend = FALSE, add = TRUE, main = "Depth-Based Suitability")
legend("bottomright", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")

par(mfrow = c(1, 1))  # reset

```

```{r}
# Plot background bathymetry with transparency
plot(depth_raster, col = gray_transparent, legend = FALSE,
     main = "Wind Power Density", axes = FALSE)

# Add wind power data
plot(wind_suitability, col = hcl.colors(50, "YlOrRd", rev = TRUE),
     add = TRUE, legend = FALSE)

# Add legend manually
legend_vals <- pretty(range(values(wind_suitability), na.rm = TRUE), 5)
legend_colors <- hcl.colors(length(legend_vals) - 1, "YlOrRd", rev = TRUE)
legend("right", legend = paste(legend_vals[-length(legend_vals)], "-", legend_vals[-1], "W/m²"),
       fill = legend_colors, title = "Wind Power", bty = "n", cex = 0.8)

```

```{r}
# Plot background again
plot(depth_raster, col = gray_transparent, legend = FALSE,
     main = "Depth Zones", axes = FALSE)

# Add zones
plot(zones3, col = zone_colors, legend = FALSE, add = TRUE)

# Add legend clearly in-frame
legend("right", legend = zone_labels, fill = zone_colors,
       title = "Depth Zones", bty = "n", cex = 0.8)

```

# Exporting

```{r}
png("output/wind_power_map.png", width = 1600, height = 1200, res = 200)

plot(depth_raster, col = gray_transparent, legend = FALSE,
     main = "Wind Power Density", axes = FALSE)

plot(wind_suitability, col = hcl.colors(50, "YlOrRd", rev = TRUE),
     add = TRUE, legend = FALSE)

legend_vals <- pretty(range(values(wind_suitability), na.rm = TRUE), 5)
legend_colors <- hcl.colors(length(legend_vals) - 1, "YlOrRd", rev = TRUE)
legend("right", legend = paste(legend_vals[-length(legend_vals)], "-", legend_vals[-1], "W/m²"),
       fill = legend_colors, title = "Wind Power", bty = "n", cex = 0.8)

dev.off()

```
```{r}
png("output/depth_zones_map.png", width = 1600, height = 1200, res = 200)

plot(depth_raster, col = gray_transparent, legend = FALSE,
     main = "Depth Zones", axes = FALSE)

plot(zones3, col = zone_colors, legend = FALSE, add = TRUE)

legend("right", legend = zone_labels, fill = zone_colors,
       title = "Depth Zones", bty = "n", cex = 0.8)

dev.off()

```

# Halfway-result, for report

```{r}
plot(depth_raster, col = gray_transparent, legend = FALSE,
     main = "Depth Zones, excluding Natura 2000", axes = FALSE)

plot(depth_zones_clean,
     col = zone_colors,
     legend = FALSE,
     add = TRUE,
     main = "Depth-Based Offshore Zones (Excluding Natura 2000)")

legend("topright", legend = zone_labels, fill = zone_colors, title = "Depth Zones", bty = "n")
```


```{r}
png("output/depth_zones_excl_natura2000.png", width = 1600, height = 1200, res = 200)

# Plot with transparent bathymetry as background
plot(depth_raster, col = gray_transparent, legend = FALSE,
     main = "Depth Zones, excluding Natura 2000", axes = FALSE)

# Overlay the cleaned zones
plot(depth_zones_clean,
     col = zone_colors,
     legend = FALSE,
     add = TRUE)

# Add legend
legend("right", legend = zone_labels, fill = zone_colors,
       title = "Depth Zones", bty = "n", cex = 0.8)

# Close the image file
dev.off()

```


